title: "SeSAMe User Guide"
shorttitle: "sesame guide"
package: sesame
output: rmarkdown::html_vignette
fig_width: 8
fig_height: 6
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{"0. SeSAMe User Guide"}
  %\VignetteEncoding{UTF-8} 
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(sesame)
library(dplyr)
options(rmarkdown.html_vignette.check_title = FALSE)
```

# SesameData

```{r, eval=FALSE}
idat_dir <- system.file("extdata/", package = "sesameData")
pfx = searchIDATprefixes(idat_dir)
> pfx
}
                                                                                4207113116_A 
"/Users/elaineficarra/.Renv/versions/4.1.dev/lib/R/library/sesameData/extdata//4207113116_A" 
                                                                                4207113116_B 
"/Users/elaineficarra/.Renv/versions/4.1.dev/lib/R/library/sesameData/extdata//4207113116_B"

# deIdentify 

```{r, eval=FALSE}
deIdentify("/Users/elaineficarra/.Renv/versions/4.1.dev/lib/R/library/sesameData/extdata/4207113116_A_Grn.idat")
deIdentify("/Users/elaineficarra/.Renv/versions/4.1.dev/lib/R/library/sesameData/extdata/4207113116_A_Red.idat")
}
