title: "de-Identifier"
shorttitle: "deIdentify
package: sesame
output: rmarkdown::html_vignette
fig_width: 8
fig_height: 6
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{"0. de-Identifier"}
  %\VignetteEncoding{UTF-8} 
---

De-Identify IDATs by masking SNP probe intensity mean by zero
Will need R package seSAMe

```{r message=FALSE, warning=FALSE, include=FALSE}
library(sesame)
library(dplyr)
options(rmarkdown.html_vignette.check_title = FALSE)
```

# sesameData

Can find high-quality DNA methylation data on more than 10,000 human samples with the HM450 platform

```{r, eval=FALSE}
idat_dir <- system.file("extdata/", package = "sesameData")
pfx = searchIDATprefixes(idat_dir)
pfx
```
                                                   
# deIdentify

DNA methylation readouts from these arrays are usually expressed in β values

```{r, eval=FALSE}

betas = getBetas(readIDATpair(pfx[3]))
head(betas[grep('^rs', names(betas))])

deIdentify(system.file(
     "extdata", "4207113116_A_Grn.idat", package = "sesameData"),
      out_path='test_Grn.idat', randomize = TRUE)
deIdentify(system.file(
     "extdata", "4207113116_A_Red.idat", package = "sesameData"),
       out_path='test_Red.idat', randomize = TRUE)
betas2 = getBetas(readIDATpair('test'))
head(betas2[grep('^rs', names(betas2))])

Whether a signal is considered real is determined by a detection P-value, representing the probability of a detected signal being background fluorescence.

}
