---
title: "Working with Mouse (MM285) Array"
package: sesame
output: rmarkdown::html_vignette
fig_width: 8
fig_height: 6
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{5. Mouse Array}
  %\VignetteEncoding{UTF-8}
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(sesame)
library(dplyr)
options(rmarkdown.html_vignette.check_title = FALSE)
```

## Load IDATs, preprocessing and masking

### Read IDATs

SeSAMe provides extensive native support for the Illumina mouse array
(referred to as the MM285 array).
The MM285 contains ~285,000 probes covering over 20 design categories including
gene promoters, enhancers, CpGs in synteny to human EPIC array as well as other
biology. This documents describe the procedure to process the MM285 array.

Let's download an example mouse array IDAT
```{r eval=FALSE}
res_grn = sesameDataDownload("204637490002_R05C01_Grn.idat")
res_red = sesameDataDownload("204637490002_R05C01_Red.idat")
pfx = sprintf("%s/204637490002_R05C01", res_red$dest_dir)
```

To load IDAT into `SigSet`, one needs the readIDATpair function,
```{r eval=FALSE}
sset = readIDATpair(pfx)
```

The default openSesame pipeline works for the mouse array
```{r eval=FALSE}
openSesame(idat_dir)
```

### Preprocessing

Let's load a pre-built `SigSet` object
```{r include=FALSE}
sset = sesameDataGet('MM285.1.NOD.FrontalLobe')
```

Preprocess the sigset to produce beta values.
The standard `noob`, `dyeBiasCorrTypeINorm` works as expected:
```{r}
sset_normalized = sset %>%
                  noob %>%
                  dyeBiasCorrTypeINorm
```
Retrieve beta values using the following commands
```{r}
betas = sset_normalized %>%
        qualityMask %>%
        detectionMask %>%
        getBetas
```

By default the repeat and suboptimally designed probes are masked by `NA`.
Starting from mouse array, the suboptimally designed probes take a new
probe ID prefix ("uk") instead of the "cg"/"ch"/"rs" typically seen in
the human array.
```{r}
sum(is.na(betas))
head(betas[grep('uk', names(betas))])
```

To use these probes, one skip qualityMask:
```{r}
betas = sset_normalized %>%
        detectionMask %>%
        getBetas
sum(is.na(betas))
head(betas[grep('uk', names(betas))])
```

Note that probes can still be masked because of insignificant detection p-value
One can completely turn off masking by skipping that
```{r}
betas = sset_normalized %>% getBetas
sum(is.na(betas))
```
or use `mask=FALSE` in the `getBetas` function.
```{r}
betas = sset_normalized %>%
        qualityMask %>%
        detectionMask %>%
        getBetas(mask = FALSE)
sum(is.na(betas))
```

## Visualize mouse array betas
```{r message=FALSE}
betas = sesameDataGet("MM285.10.tissue")$betas
visualizeGene("Igf2", betas = betas, platform="MM285", refversion = "mm10")
```

