---
title: "Working with Mouse (MM285) Array"
date: "`r BiocStyle::doc_date()`"
package: sesame
output: BiocStyle::html_document
fig_width: 8
fig_height: 6
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{6. Mouse Array}
  %\VignetteEncoding{UTF-8}
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(sesame)
library(wheatmap)
library(dplyr)
options(rmarkdown.html_vignette.check_title = FALSE)
```

## Load IDATs, preprocessing and masking

### Cache mouse array annotation data

To begin, we need to retrieve mouse annotation data from ExperimentHub.
This only needs to be done once per sesame installation.
```{r}
sesameDataCache("MM285")
```

### Read IDATs

SeSAMe provides extensive native support for the Illumina mouse array
(referred to as the MM285 array).
The MM285 contains ~285,000 probes covering over 20 design categories including
gene promoters, enhancers, CpGs in synteny to human EPIC array as well as other
biology. This documents describe the procedure to process the MM285 array.

Let's download an example mouse array IDAT
```{r eval=FALSE}
res_grn = sesameDataDownload("204637490002_R05C01_Grn.idat")
res_red = sesameDataDownload("204637490002_R05C01_Red.idat")
pfx = sprintf("%s/204637490002_R05C01", res_red$dest_dir)
```

To load IDAT into `SigSet`, one needs the readIDATpair function,
```{r eval=FALSE}
sset = readIDATpair(pfx)
```

The default openSesame pipeline works for the mouse array
```{r eval=FALSE}
openSesame(idat_dir)
```

### Preprocessing

Let's load a pre-built `SigSet` object
```{r include=FALSE}
sset = sesameDataGet('MM285.1.NOD.FrontalLobe')
```

Preprocess the sigset to produce beta values.
The standard `noob`, `dyeBiasCorrTypeINorm` works as expected:
```{r}
sset_normalized = sset %>%
                  noob %>%
                  dyeBiasCorrTypeINorm
```
Retrieve beta values using the following commands
```{r}
betas = sset_normalized %>%
        qualityMask %>%
        detectionMask %>%
        getBetas
```

By default the repeat and suboptimally designed probes are masked by `NA`.
Starting from mouse array, the suboptimally designed probes take a new
probe ID prefix ("uk") instead of the "cg"/"ch"/"rs" typically seen in
the human array.
```{r}
sum(is.na(betas))
head(betas[grep('uk', names(betas))])
```

To use these probes, one skip qualityMask:
```{r}
betas = sset_normalized %>%
        detectionMask %>%
        getBetas
sum(is.na(betas))
head(betas[grep('uk', names(betas))])
```

Note that probes can still be masked because of insignificant detection p-value
One can completely turn off masking by skipping that
```{r}
betas = sset_normalized %>% getBetas
sum(is.na(betas))
```
or use `mask=FALSE` in the `getBetas` function.
```{r}
betas = sset_normalized %>%
        qualityMask %>%
        detectionMask %>%
        getBetas(mask = FALSE)
sum(is.na(betas))
```

## Visualize mouse array betas
```{r message=FALSE}
betas = sesameDataGet("MM285.10.tissue")$betas
visualizeGene("Igf2", betas = betas, platform="MM285", refversion = "mm10")
```

## Infer Strain Information

Let's load a pre-built `SigSet` object from SeSAMeData
```{r}
sset <- sesameDataGet('MM285.1.NOD.FrontalLobe')
```

Calculate beta values using the following commands.
```{r}
betas <- sset %>%
         noob %>%
         dyeBiasCorrTypeINorm %>%
         getBetas
```

Convert the beta values to Variant Allele Frequencies.  
It should be noted that since variant allele frequency is not always measured in green for Infinium-II and M-allele for Infinium-I, one needs to flip the beta values for some probes to calculate variant allele frequency.
```{r}
vafs <- betaToAF(betas)
```

Infer strain information for mouse array. This will return a list containing the best guess, p-value of the best guess, and probabilities of all strains.
```{r}
strain <- inferStrain(vafs)
strain$pval
```
Let's visualize the probabilities of other strains.
```{r fig.width=6, fig.height=5}
library(ggplot2)
df <- data.frame(strain=names(strain$probs), probs=strain$probs)
ggplot(data = df,  aes(x = strain, y = log(probs))) +
  geom_bar(stat = "identity", color="gray") +
  ggtitle("strain probabilities") +
  scale_x_discrete(position = "top") +
  theme(axis.text.x = element_text(angle = 90), legend.position = "none")
```

## Contrast Data with Tissue References

Let's load beta values from SeSAMeData
```{r}
betas <- sesameDataGet("MM285.10.tissue")$betas[,1:2]
```

Compare mouse array data with mouse tissue references. This will return a grid object that contrasts the traget sample with pre-build mouse tissue reference.
```{r fig.width=6, fig.height=5}
compareMouseTissueReference(betas)
```

## Infer Mouse Age

Let's load beta values from SeSAMeData
```{r}
betas <- sesameDataGet('MM285.10.tissue')$betas
```

The age of the mouse can be predicted using the `predictMouseAgeInMonth` function. This looks for overlapping probes and estimates age using an aging model built from 347 MM285 probes. The function returns a numeric output of age in months.  
The model is most accurate with SeSAMe preprocessing.
Here's an example.
```{r}
predictMouseAgeInMonth(betas[,1])
```
This indicates thaat this mouse is approximately 1.41 months old.