---
title: "Working with Non-human Array"
date: "`r BiocStyle::doc_date()`"
package: sesame
output: BiocStyle::html_document
fig_width: 6
fig_height: 6
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{2. Non-human Array}
  %\VignetteEncoding{UTF-8}
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(sesame)
library(wheatmap)
library(dplyr)
options(rmarkdown.html_vignette.check_title = FALSE)
```

# Mouse MM285 Array

## Preprocessing

To begin, we need to retrieve mouse annotation data from ExperimentHub.
This only needs to be done once per sesame installation.
```{r echo=FALSE, message=FALSE}
sesameDataCache("MM285")
```

SeSAMe provides extensive native support for the Illumina mouse array
(referred to as the MM285 array).
The MM285 contains ~285,000 probes covering over 20 design categories including
gene promoters, enhancers, CpGs in synteny to human EPIC array as well as other
biology. This documents describe the procedure to process the MM285 array.

Let's download an example mouse array IDAT
```{r eval=FALSE}
res_grn = sesameDataDownload("204637490002_R05C01_Grn.idat")
res_red = sesameDataDownload("204637490002_R05C01_Red.idat")
pfx = sprintf("%s/204637490002_R05C01", res_red$dest_dir)
```

To load IDAT into `SigSet`, one needs the readIDATpair function,
```{r eval=FALSE}
sdf = readIDATpair(pfx)
```

The default openSesame pipeline works for the mouse array
```{r eval=FALSE}
openSesame(idat_dir)
```

Let's load a pre-built `SigSet` object
```{r include=FALSE}
sdf = sesameDataGet('MM285.1.SigDF')
```

Preprocess the sigset to produce beta values.
The standard `noob`, `dyeBiasCorrTypeINorm` works as expected:
```{r}
sdf_normalized = sdf %>%
                 qualityMask %>%
                 pOOBAH %>%
                 noob %>%
                 dyeBiasCorrTypeINorm
```
Retrieve beta values using the following commands
```{r}
betas = getBetas(sdf_normalized)
```

By default the repeat and suboptimally designed probes are masked by `NA`.
Starting from mouse array, the suboptimally designed probes take a new
probe ID prefix ("uk") instead of the "cg"/"ch"/"rs" typically seen in
the human array.
```{r}
sum(is.na(betas))
head(betas[grep('uk', names(betas))])
```

To use these probes, one skip qualityMask and explicitly perform masking based
on detection p-values only:
```{r}
betas = sdf_normalized %>%
        setMask(pOOBAH(qualityMask(sdf), return.pval=TRUE)>0.05) %>%
        getBetas
sum(is.na(betas))
head(betas[grep('uk', names(betas))])
```
Not that we still use qualityMask for calculating p-values. In this example,
probes are only masked because of insignificant detection p-value
One can completely turn off all masking by toggling off the `mask` option in
`getBetas`:
```{r}
betas = getBetas(sdf_normalized, mask = FALSE)
sum(is.na(betas))
```
or reset the mask using `resetMask` function
```{r}
betas = getBetas(resetMask(sdf_normalized))
sum(is.na(betas))
```

## Track View
```{r message=FALSE}
betas = sesameDataGet("MM285.10.tissue")$betas
visualizeGene("Igf2", betas = betas, platform="MM285", refversion = "mm10")
```

## Species, Strain Inference

Let's load a pre-built `SigSet` object from SeSAMeData
```{r}
sdf <- sesameDataGet("MM285.1.SigDF")
```

Infer species (obviously this is a mouse, but this is supposed to work on rat,
human etc). Currently this feature supports both the mouse array and the mammal
array.
```{r}
inferSpecies(sdf, return.species = TRUE)
```

Calculate beta values using the following commands.
```{r}
betas <- sdf %>%
         noob %>%
         dyeBiasCorrTypeINorm %>%
         getBetas
```

Infer strain information for mouse array. This will return a list containing
the best guess, p-value of the best guess, and probabilities of all strains.
Internally, the function converts the beta values to variant allele
frequencies. It should be noted that since variant allele frequency is not
always measured as the M-allele, one needs to flip the beta values for some
probes to calculate variant allele frequency.

```{r}
inferStrain(sdf, return.strain = TRUE)
p <- inferStrain(sdf, return.probability = TRUE)
```
Let's visualize the probabilities of other strains.
```{r}
library(ggplot2)
df <- data.frame(strain=names(p), probs=p)
ggplot(data = df,  aes(x = strain, y = probs)) +
  geom_bar(stat = "identity", color="gray") +
  ggtitle("Strain Probabilities") +
  ylab("Probability") + xlab("") +
  scale_x_discrete(position = "top") +
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=0),
    legend.position = "none")
```

## Tissue Type Inference

Let's load beta values from SeSAMeData
```{r}
betas <- sesameDataGet("MM285.10.tissue")$betas[,1:2]
```

Compare mouse array data with mouse tissue references. This will return a grid
object that contrasts the traget sample with pre-build mouse tissue reference.
```{r}
options(sesameData_use_alternative=TRUE)
compareMouseTissueReference(betas)
```

## Sex Inference

Let's load beta values from sesameData
```{r}
sdf = sesameDataGet("MM285.1.SigDF")
```

Sex inference can take both the raw signal in SigDF or beta value vector
```{r}
inferSex(sdf)
```

## Age Inference

Let's load beta values from SeSAMeData
```{r}
betas <- sesameDataGet('MM285.10.tissue')$betas
```

The age of the mouse can be predicted using the `predictMouseAgeInMonth`
function. This looks for overlapping probes and estimates age using an aging
model built from 347 MM285 probes. The function returns a numeric output of
age in months. The model is most accurate with SeSAMe preprocessing.
Here's an example.
```{r}
predictMouseAgeInMonth(betas[,1])
```
This indicates thaat this mouse is approximately 1.41 months old.

## Human-Mouse Mixture

UNDER CONSTRUCTION

# Horvath Mammal40 Array

SeSAMe supports Mammal 40 array natively.

```{r echo=FALSE, message=FALSE}
sesameDataCache("Mammal40")
```

```{r}
res_grn = sesameDataDownload("GSM4411982_Grn.idat.gz")
res_red = sesameDataDownload("GSM4411982_Red.idat.gz")
sdf = readIDATpair(sprintf("%s/GSM4411982", res_red$dest_dir))
```

Preprocess the sigset to produce beta values.
The standard `noob`, `dyeBiasCorrTypeINorm` works as expected:
```{r}
sdf_normalized = dyeBiasCorrTypeINorm(noob(pOOBAH(qualityMask(sdf))))
```
Retrieve beta values using the following commands
```{r}
betas = getBetas(sdf_normalized)
head(betas)
```

# Session Info

```{r}
sessionInfo()
```