---
title: "Quality Controls"
date: "`r BiocStyle::doc_date()`"
package: sesame
output: BiocStyle::html_document
fig_width: 8
fig_height: 6
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{1. Quality Controls}
  %\VignetteEncoding{UTF-8}
---

# Generation of Quality Controls

SeSAMe provides a set of quality control steps.
Be sure to pre-cache HM450 annotation data from ExperimentHub.
This only needs to be done once per sesame installation.
```{r, echo = FALSE, message = FALSE}
library(sesame)
library(FlowSorted.Blood.450k)
options(rmarkdown.html_vignette.check_title = FALSE)
sesameDataCache("HM450")
sesameDataCache("EPIC")

ssets <- RGChannelSetToSigSets(FlowSorted.Blood.450k[,1:10])
```

The SeSAMe QC function returns an `sesameQC` object which can be
directly printed onto the screen.
```{r}
sesameQC(ssets[[1]])
```

The `sesameQC` object can be coerced into data.frame and linked
using the following code
```{r}
qc10 <- do.call(rbind, lapply(ssets, function(x)
    as.data.frame(sesameQC(x))))
qc10$sample_name <- names(ssets)

qc10[,c('mean_beta_cg','frac_meth_cg','frac_unmeth_cg','sex','age')]
```

# Background

The background level is given by `mean_oob_grn` and `mean_oob_red`
```{r}
library(ggplot2)
ggplot(qc10,
    aes(x = mean_oob_grn, y= mean_oob_red, label = sample_name)) +
    geom_point() + geom_text(hjust = -0.1, vjust = 0.1) +
    geom_abline(intercept = 0, slope = 1, linetype = 'dotted') +
    xlab('Green Background') + ylab('Red Background') +
    xlim(c(500,1200)) + ylim(c(500,1200))
```

# Mean Intensity

The mean {M,U} intensity can be reached by `mean_intensity`.
Similarly, the mean M+U intensity can be reached by
`mean_intensity_total`. Low intensities are symptomatic of low
input or poor hybridization.

```{r warning = FALSE}
library(wheatmap)
p1 <- ggplot(qc10) +
    geom_bar(aes(sample_name, mean_intensity), stat='identity') +
    xlab('Sample Name') + ylab('Mean Intensity') +
    ylim(0,18000) +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))
p2 <- ggplot(qc10) +
    geom_bar(aes(sample_name, mean_intensity_total), stat='identity') +
    xlab('Sample Name') + ylab('Mean M+U Intensity') +
    ylim(0,18000) +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))
WGG(p1) + WGG(p2, RightOf())
```

# Fraction of color channel switch

The fraction of color channel switch can be found in 
`InfI_switch_G2R` and `InfI_switch_R2G`. These numbers are 
symptomatic of how Infinium I probes are affected by SNP-induced
color channel switching.
```{r}
ggplot(qc10) +
    geom_point(aes(InfI_switch_G2R, InfI_switch_R2G))
```

# Fraction of NA

The fraction of NAs are signs of masking due to variety of reasons
including failed detection, high background, putative low quality
probes etc. This number can be reached in `frac_na_cg` and 
`num_na_cg` (the cg stands for CpG probes, so we also have
`num_na_ch` and `num_na_rs`)
```{r}
p1 <- ggplot(qc10) +
    geom_bar(aes(sample_name, num_na_cg), stat='identity') +
    xlab('Sample Name') + ylab('Number of NAs') +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))
p2 <- ggplot(qc10) +
    geom_bar(aes(sample_name, frac_na_cg), stat='identity') +
    xlab('Sample Name') + ylab('Fraction of NAs (%)') +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))
WGG(p1) + WGG(p2, RightOf())
```

# Quality Ranking

Sesame provide convenient function to compare your sample with
public data sets processed with the same pipeline. All you need
is a raw SigSet.
```{r}
sset <- sesameDataGet('EPIC.1.LNCaP')$sset
qualityRank(sset)
```

# Output explicit and Infinium-I-derived SNP to VCF

```{r}
sset <- sesameDataGet('EPIC.1.LNCaP')$sset

annoS <- sesameDataPullVariantAnno_SNP('EPIC','hg19')
annoI <- sesameDataPullVariantAnno_InfiniumI('EPIC','hg19')

## output to console
head(formatVCF(sset, annoS=annoS, annoI=annoI))
```

One can output to actual VCF file with a header by `formatVCF(sset,
vcf=path_to_vcf)`.
