---
title: "Quality Control"
date: "`r BiocStyle::doc_date()`"
package: sesame
output: BiocStyle::html_document
fig_width: 6
fig_height: 5
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{1. Quality Control}
  %\VignetteEncoding{UTF-8}
---

```{r message=FALSE, results="hide"}
library(sesame)
sesameDataCache()
```

# Calculate quality metrics

The central function to calculate the quality metrics is `sesameQC_calcStats`.
SeSAMe divides sample quality metrics into multiple groups. Each group of
quality metrics can be generated using a dedicated function. All these
functions can be accessed through the `sesameQC_calcStats`
function. `sesameQC_calcStats()` calculates QC stats of one SigDF and returns a
single object of the class `sesameQC`.

```{r qc1}
sdf <- sesameDataGet('EPIC.1.SigDF')
qc <- sesameQC_calcStats(sdf, c("dyeBias", "detection"))
qc
```

The `sesameQC_calcStats` function returns an S4 `sesameQC` object. The choice
of QC metrics depends on the 2nd argument of the `sesameQC_calcStats`
function. This argument is optional and can take one or a list of the following
keys. For example, "intensity" generates signal intensity related quality
metrics. When the 2nd argument is not given, all stats will be calculated by
default. The following demonstrates the use of these functions. We consider
signal detection the most important QC metric.

```{r echo=FALSE}
library(knitr)
kable(data.frame(
    "Short Key" = c(
        "detection",
        "numProbes",
        "intensity",
        "channel",
        "dyeBias",
        "betas"),
    "Description" = c(
        "Signal Detection",
        "Number of Probes",
        "Signal Intensity",
        "Color Channel",
        "Dye Bias",
        "Beta Value")))
```

One can calculate specific group(s) of metrics by giving the 2nd argument: e.g.,

One can retrieve the actual stats using the sesameQC_getStats

```{r qc2}
sesameQC_getStats(qc, "frac_dt")
```

One can combine multiple sesameQC into a data frame:

```{r eval=FALSE}
sdfs <- sesameDataGet("EPIC.5.SigDF.normal")
df <- do.call(rbind, lapply(sdfs, function(sdf)
   as.data.frame(sesameQC_calcStats(sdf, "detection"))))
```

# Rank quality metrics

```{r qc4, echo=FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
```

SeSAMe features functions that compare your sample with public data sets. The
`sesameQC_rankStats()` function rank the input `sesameQC` object with
`sesameQC` calculated from public datasets. It shows the rank percentage of the
input sample as well as the number of datasets compared.

```{r qc5}
sdf <- sesameDataGet('EPIC.1.SigDF')
qc <- sesameQC_calcStats(sdf, "intensity")
qc
sesameQC_rankStats(qc, platform="EPIC")
```

# Quality control plots

SeSAMe provides functions to create QC plots. Some functions takes sesameQC as
input while others directly plots the SigDF objects. For example, the
`sesameQC_plotBar` function takes a list of sesameQC objects and creates bar
plot for each metric calculated. More about quality control plots can be found
in [Supplemental
Vignette](https://zhou-lab.github.io/sesame/supplemental.html).

# Session Info

```{r}
sessionInfo()
```