---
title: "Quality Control"
date: "`r BiocStyle::doc_date()`"
package: sesame
output: BiocStyle::html_document
fig_width: 6
fig_height: 6
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{1. Quality Control}
  %\VignetteEncoding{UTF-8}
---

# Quality Ranking

```{r echo=FALSE, message=FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
library(sesame)
sesameDataCacheAll()
```

Probe detection success rate is the main QC metric for an array
experiment.
Sesame provide convenient function to compare your sample with
public data sets processed with the same pipeline. All you need
is a raw SigSet.
```{r}
sdf <- sesameDataGet('EPIC.1.SigDF')
sesameQC_rankDetectionSuccess(sdf)
sesameQC_rankMeanIntensity(sdf)
```

## Signal Background

The background level is given by `mean_oob_grn` and `mean_oob_red`
```{r message = FALSE}
sdfs = sesameDataGet("EPIC.5.SigDF.normal")
```

```{r}
library(ggplot2)
qc5_intens = sesameQC_calcStats(sdfs, sesameQC_calcStats_intens)
ggplot(qc5_intens,
    aes(x = mean_oob_grn, y= mean_oob_red, label = sample_name)) +
    geom_point() + geom_text(hjust = -0.1, vjust = 0.1) +
    geom_abline(intercept = 0, slope = 1, linetype = 'dotted') +
    xlab('Green Background') + ylab('Red Background') +
    xlim(c(200,600)) + ylim(c(200,600))
```

## Mean Intensity

The mean {M,U} intensity can be reached by `mean_intensity`.
Similarly, the mean M+U intensity can be reached by
`mean_intensity_total`. Low intensities are symptomatic of low
input or poor hybridization.

```{r warning = FALSE}
library(wheatmap)
p1 <- ggplot(qc5_intens) +
    geom_bar(aes(sample_name, mean_intensity), stat='identity') +
    xlab('Sample') + ylab('Mean Intensity') +
    ylim(0,18000) +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))
p2 <- ggplot(qc5_intens) +
    geom_bar(aes(sample_name, mean_intensity_total), stat='identity') +
    xlab('Sample') + ylab('Mean M+U Intensity') +
    ylim(0,18000) +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))
WGG(p1) + WGG(p2, RightOf())
```

## Probe Success Rate

The fraction of NAs are signs of masking due to variety of reasons
including failed detection, high background, putative low quality
probes etc. This number can be reached in `frac_na_cg` and 
`num_na_cg` (the cg stands for CpG probes, so we also have
`num_na_ch` and `num_na_rs`)
```{r}
qc5_detection = sesameQC_calcStats(sdfs, sesameQC_calcStats_detection)
p1 <- ggplot(qc5_detection) +
    geom_bar(aes(sample_name, num_na_cg), stat='identity') +
    xlab('Sample') + ylab('Number of NAs') +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))
p2 <- ggplot(qc5_detection) +
    geom_bar(aes(sample_name, frac_na_cg*100), stat='identity') +
    xlab('Sample') + ylab('Fraction of NAs (%)') +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))
WGG(p1) + WGG(p2, RightOf())
```

## Color Channel Switch

The fraction of color channel switch can be found in 
`InfI_switch_G2R` and `InfI_switch_R2G`. These numbers are 
symptomatic of how Infinium I probes are affected by SNP-induced
color channel switching.
```{r fig.height=3, fig.width=8}
qc5_channel = sesameQC_calcStats(sdfs, sesameQC_calcStats_channel)
p1 = ggplot(qc5_channel) +
   geom_bar(aes(sample_name, InfI_switch_R2G),
   fill="seagreen", stat="identity") +
   xlab("Sample") + ylab("Infinium-I R>G switch")
p2 = ggplot(qc5_channel) +
   geom_bar(aes(sample_name, InfI_switch_G2R), fill="red", stat="identity") +
   xlab("Sample") + ylab("Infinium-I G>R switch")
WGG(p1) + WGG(p2, RightOf())
```

## Bisulfite Conversion

Infinium platforms are intrinsically robust to incomplete bisulfite conversion
as non-converted probes would fail to hybridize to the target. Residual
incomplete bisulfite conversion can be quantified using GCT score based on
C/T-extension probes. Details of this method can be found in
[Zhou et al. 2017](https://academic.oup.com/nar/article/45/4/e22/2290930).
The closer the score to 1.0, the more complete the bisulfite conversion.
```{r}
sdf <- sesameDataGet('EPIC.1.SigDF')
bisConversionControl(sdf)
```

# Extract Genotypes

SeSAMe can output explicit and Infinium-I-derived SNP to VCF.
This information can be used to identify sample swaps.

```{r}
annoS = sesameDataGetAnno("EPIC/EPIC.hg19.snp_overlap_b151.rds")
annoI = sesameDataGetAnno("EPIC/EPIC.hg19.typeI_overlap_b151.rds")
head(formatVCF(sdf, annoS=annoS, annoI=annoI))
```

One can output to actual VCF file with a header by `formatVCF(sdf,
vcf=path_to_vcf)`.

# Other Quality Metrics

SeSAMe provides a set of quality control functions. These include
`sesameQC_calcStats_intens`, `sesameQC_calcStats_numProbes`,
`sesameQC_calcStats_channel`, `sesameQC_calcStats_detection`,
and `sesameQC_calcStats_betas`, and sesameQC_calcStats, which calculates
all the qc stats. Each sesameQC_calcStats_ function returns a data
frame. Using `verbose=TRUE` print readable information to the console.
```{r}
sdf = sesameDataGet("EPIC.1.SigDF")
sesameQC_calcStats_channel(sdfs[[1]])
```

# Session Info

```{r}
sessionInfo()
```