---
title: "Quality Control"
date: "`r BiocStyle::doc_date()`"
package: sesame
output: BiocStyle::html_document
fig_width: 6
fig_height: 6
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{1. Quality Control}
  %\VignetteEncoding{UTF-8}
---

```{r message=FALSE, results="hide"}
library(sesame)
sesameDataCacheAll()
```

# Rank quality

```{r echo=FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
```

Probe detection success rate is the main QC metric for an array
experiment.
Sesame provide convenient function to compare your sample with
public data sets. One will need a raw SigSet. The function returns
the fraction of public data sets the target data has beaten in
detection success rate.

```{r}
sdf <- sesameDataGet('EPIC.1.SigDF')
sesameQC_calcStats_rank(sdf)
```

One can calculate the quality rank
```{r}
sdfs <- sesameDataGet("EPIC.5.SigDF.normal")
lapply(sdfs, sesameQC_rankStats)
```

# Calculate quality metrics

`sesameQC_calcStats` calculates QC stats from one or a list of SigDFs. Instead
of returning a single `sesameQC` object, a data.frame combining multiple
samples will be returned.

```{r}
sesameQC_calcStats(sdfs)
```

The 2nd argument of the function is optional and specifies one of the
`sesameQC_calcStats_*` function listed below (the "Available quality metrics"
section).

By default, sesame calculates detection metrics

```{r}
sesameQC_calcStats(sdfs, sesameQC_calcStats_detection)
```

But one can choose to calculate other metrics or a combination of metrics:

```{r}
sesameQC_calcStats(sdfs,
    c(sesameQC_calcStats_dyeBias, sesameQC_calcStats_detection))
```

# Available quality metrics

SeSAMe divides sample quality metrics into multiple groups. Each group of
quality metrics can be generated using a dedicated function. These functions
share a `sesameQC_calcStats_` prefix.

For example, `sesameQC_calcStats_intens`, generates signal intensity related
quality metrics. The following demonstrates the use of these functions.

## Signal intensity

```{r}
sesameQC_calcStats_intens(sdf)
```

## Signal detection

```{r}
sesameQC_calcStats_detection(sdf)
```

## Dye bias

```{r}
sesameQC_calcStats_dyeBias(sdf)
```

## Probe count

```{r}
sesameQC_calcStats_numProbes(sdf)
```

## Color channel switch

```{r}
sesameQC_calcStats_channel(sdf)
```

## Beta value distribution

```{r}
sesameQC_calcStats_betas(sdf)
```

# Example quality control plots

```{r result="hide", message=FALSE}
library(ggplot2)
library(wheatmap)
```

## Probe success rate

The fraction of detection failures are signs of masking due to variety of
reasons including failed detection, high background, putative low quality
probes etc. This number can be reached in `frac_na_cg` and `num_na_cg` (the cg
stands for CpG probes, so we also have `num_na_ch` and `num_na_rs`)

```{r fig.height=3, fig.width=8}
sesameQC_plotBarDetection(sdfs)
```

## Signal intensity

Low intensities are symptomatic of low input or poor hybridization.
The following compares mean intensity among samples.

```{r fig.height=3, fig.width=8}
sesameQC_plotBarIntens(sdfs)
```

## Color channel switch

The fraction of color channel switch can be found in 
`InfI_switch_G2R` and `InfI_switch_R2G`. These numbers are 
symptomatic of how Infinium I probes are affected by SNP-induced
color channel switching.

```{r fig.height=3, fig.width=8}
sesameQC_plotBarChannel(sdfs) 
```

## Dye bias Q-Q plot

Dye bias is shown by an off-diagonal q-q plot of the red (x-axis) and green
signal (y-axis).

```{r fig.height=5, fig.width=5, fig.align="center"}
sesameQC_plotRedGrnQQ(sdf)
```

## Intensity-beta plot

Beta value is more influenced by signal background for probes with low signal
intensities. The following plot shows this dependency and the extent of probes
with low signal intensity.

```{r fig.height=5, fig.width=5, fig.align="center"}
sesameQC_plotIntensVsBetas(sdf)
```

# Extract genotypes

sesameDataGetAnno("")

SeSAMe can output explicit and Infinium-I-derived SNPs to create a VCF file.
This information can be used to identify sample swaps.

```{r message=FALSE, results="hide"}
annoS <- sesameDataGetAnno("EPIC/EPIC.hg19.snp_overlap_b151.rds")
annoI <- sesameDataGetAnno("EPIC/EPIC.hg19.typeI_overlap_b151.rds")
```

```{r}
head(formatVCF(sdf, annoS=annoS, annoI=annoI))
```

One can output to actual VCF file with a header by `formatVCF(sdf,
vcf=path_to_vcf)`.

# Bisulfite conversion

Infinium platforms are intrinsically robust to incomplete bisulfite conversion
as non-converted probes would fail to hybridize to the target. Residual
incomplete bisulfite conversion can be quantified using GCT score based on
C/T-extension probes. Details of this method can be found in
[Zhou et al. 2017](https://academic.oup.com/nar/article/45/4/e22/2290930).
The closer the score to 1.0, the more complete the bisulfite conversion.

```{r}
sdf <- sesameDataGet('EPIC.1.SigDF')
bisConversionControl(sdf)
```

# Session Info

```{r}
sessionInfo()
```